Natural Selection in the 20th Century

### 2020-11-12 why-negative-selection.Rmd
papers; notes; exploration of sex, pay, married couples

---
title: "Natural Selection in the 20th Century"
author: "David Hugh-Jones, Abdel Abdellaoui"
date: "`r Sys.Date()`"
abstract: "Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank. Selection effects appear consistent over time. The strength of natural selection on a polygenic score can be predicted by its correlation with earnings and education. Effects are larger among low-income and less educated individuals, younger parents, and people not living with a partner. Effect sizes are larger when we correct for ascertainment bias. The direction of natural selection is reversed among older parents, or  controlling for age at first live birth. These patterns can be explained by economic theories of fertility, in which higher earnings can either increase or decrease fertility."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
editor_options: 
  markdown: 
    wrap: 72
---

@@

# Introduction

@@

```{r TODO, eval = FALSE}
# Open questions/TODO

* If we weight with the GHS, and plot the means of figure 1 using this w
  weighting, most things stay the same - but the change in EA is reversed! EA
  increases over time. Could this be because the sample includes many
  relatively educated older people? If so, then mean effects are *overestimated*
  by figure 1. 
* Reweight other regressions (e.g. the omnibus regression in the appendix)?
* Look at geography, esp. in siblings regressions.
* We say we don't see change over time. But, we do see the effect of 
  controlling for AFLB changing over generations.
* Beauchamp 2016 used the HRS. He reckons it's representative. Is it? How do
  his effect sizes compare?
  - he also worries about the quality/quantity tradeoff, which might mean that
    people with fewer children have more grandchildren (than you'd expect). 
    As we have the two generations, we can test this. (Actually, there's a 
    positive correlation of about 0.1)
* Another point is that negative selection only among poor will increase 
    genetic inequality.
```

# Results

@@

```{r plot-means-over-time, fig.height = 5, fig.cap = "Mean polygenic scores by birth year in UK Biobank. Points are means for 5-year intervals. Lines are 95% confidence intervals. Green lines show a significant linear increase over time (p < 0.05/33). Red lines show a significant decrease."}

@@

pgs_over_time %>% 
      mutate(score_name = pretty_names(score_name)) %>% 
      ggplot(aes(YOB, group = score_name, colour = Change)) + 
      geom_pointrange(aes(ymin = score_lo, y = score, ymax = score_hi), 
        fatten = 2) +
      facet_wrap(~score_name, ncol = 5) +
      my_hline +
      scale_color_manual(values = c("-" = "red", "+" = "darkgreen", "o" = "black")) +
      theme(
        strip.text  = element_text(size = 6), 
        axis.text.x = element_text(size = 7),
        legend.position = "none"
      ) +
      labs(y = "PGS", x = "Birth year (5 year intervals)")

@@

Why do selection effects have *opposite* signs across older and younger
mothers at first live birth, and opposite signs when AFLB is controlled
for? A possible answer is given by the economic theory of fertility
[@becker1960economic]. According to this, increases in potential earned
income affect fertility via two opposing channels. There is an "income
effect" by which children become more affordable, like any other good.
Since childrearing has a cost in time, there is also a "substitution
effect": the opportunity cost of childrearing increases if one's market wage is
higher. The income effect would lead higher earners to have more children. The
substitution effect would lead higher earners to have fewer children. It is
often assumed that the substitution effect will dominate for lone parents, or
those in unstable relationships, since they have less opportunity to share
childcare responsibilities, while the income effect will dominate for couples
who are able to reap the gains from specialization.

@@

# Conclusion

@@

An alternative theory is that traits selected for are linked to
externalizing behaviour and low time discount rates, via the channel of
early sexual behaviour [@Mills2020.05.06.081273]. The data here provide
some support for this: scores which might plausibly be linked to
externalizing behaviour, like ADHD and age at smoking initiation, are
selected for. However, this theory is less good at explaining variation
in selection across the full range of scores, including physical measures, e.g. waist-hip ratio and BMI. Externalizing behaviour also does not explain
that selection effects change sign among older parents. Overall, we believe
that the economic theory is the leading explanation at present.

PGS which correlate with high (low) earnings and more (less) education are being
selected against (selected for). In addition, many of the phenotypes under
positive selection are linked to poor health, or would be seen as undesirable to
have by many people. For example, most people would probably prefer to have high
educational attainment, a low risk of major depressive disorder, and a low risk
of coronary artery disease, but natural selection is pushing against these
traits. Potentially, this could increase the health burden on modern societies,
but that depends on effect sizes.

Our results suggest that naïve estimates can be affected by sample ascertainment
bias. This problem may be less serious in surveys which aim to be representative
(as the UK Biobank does not). Nevertheless, not all respondents consent to the
collection of genetic data: for instance, completion rates for genotype data in
the US Health and Retirement Study  were around 80-85% [@hrs2020]. Researchers
should be aware of the risks of ascertainment when studying contemporary 
natural selection.

@@

Research on natural selection is controversial. It brings to mind eugenics,
which has been used to justify mass murder and human rights violations, in the
past and possibly the present [@mao1998chinese; @sleeboom2011genetic;
@zenz2020]. Whilst we claim scientific legitimacy for our empirical results, we
do not claim to speak with any special authority on how society should react to
them. This can only be resolved by open debate, in which scientists take part as
equals among many. We therefore choose not to discuss policy implications here.

@@

@@

\FloatBarrier
## Selection over time

Tables \@ref(tab:tbl-siblings-by-period) and
\@ref(tab:tbl-children-by-period) summarize the results. There is no
change in natural selection in the parents' generation. In the
respondents' generation, effect sizes were larger in absolute size among
the late born for five PGS: autism, cognitive ability, EA2, EA3 and
extraversion. Three PGS showed significant changes in sign: alcohol use
(positive to negative); conscientiousness (negative to positive); and
type II diabetes (negative to positive).

@@

Another way to examine the "earnings" theory of natural selection is to
compare effect sizes of PGS with their genetic correlation with
educational attainment (EA3). Since EA3 strongly predicts earnings, if
earnings drives differences in fertility, we'd expect a correlation
between the two sets of results. Figure
\@ref(fig:plot-rgs-by-effect-size) shows this is so.

>>>>

---
title: "Natural Selection in the 20th Century"
author: "David Hugh-Jones, Abdel Abdellaoui"
date: "`r Sys.Date()`"
abstract: "Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank. Selection effects appear consistent over time. The strength of natural selection on a polygenic score can be predicted by its correlation with earnings and education. Effects are larger among low-income and less educated individuals, younger parents, and people not living with a partner. Effect sizes are larger when we correct for ascertainment bias. The direction of natural selection is reversed among older parents, or  controlling for age at first live birth. These patterns can be explained by economic theories of fertility, in which higher earnings can either increase or decrease fertility."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
editor_options: 
  markdown: 
    wrap: 72
---

@@

@@

```{r NOTES-TODO, eval=FALSE}
"
# Open questions/TODO

* If we weight with the GHS, and plot the means of figure 1 using this w
  weighting, most things stay the same - but the change in EA is reversed! EA
  increases over time. Could this be because the sample includes many
  relatively educated older people? If so, then mean effects are *overestimated*
  by figure 1.
* Reweight other regressions (e.g. the omnibus regression in the appendix)?
* Look at geography, esp. in siblings regressions.
* We say we don't see change over time. But, we do see the effect of
  controlling for AFLB changing over generations.
* Beauchamp 2016 used the HRS. He reckons it's representative. Is it? How do
  his effect sizes compare?
  - he also worries about the quality/quantity tradeoff, which might mean that
    people with fewer children have more grandchildren (than you'd expect).
    As we have the two generations, we can test this. (Actually, there's a
    positive correlation of about 0.1)
* Another point is that negative selection only among poor will increase
    genetic inequality.
* When we have better definition of "partner", bring it across from the other
  project to define with_partner
* If you take "year of last job started" and see if it is after/before
  the first child, you can figure out whether mothers keep working.
  - better still, look at all jobs started and see if any are within e.g. 5 years of
  first child's birth
  - this could then be another way to get at income vs substitution effects. You'd
  predict negative selection effects among mothers who worked. Of course the
  choice to work is endogenous.
* Put mediating effects of pay and education in there (see explore-sex-vs-pay.R)
  - in a multivariate regression, most effects vanish once we control for these, including EA3

## LITERATURE NOTES

* Lundberg and Pollak JEP 2007:
    - several models where nonmarital fertility is more likely for poorer women
    - Burdett and Ermisch 2002, Willis 1999, Neal 2004
    - Unmarried parents may not be able to credibly commit to each other

* Happel et al. 1994
   - childbirth is chosen to minimize loss of earnings from interrupted work.
   - early childbirth if your skills decay a lot  while childrearing. (No point
     gaining them then losing them.) Late childbirth otherwise.
   - late childbirth if you have acquired skills before marriage.
   - Capital-market imperfections make it wise to delay childbirth (till man has
     high income to cope while wife is at home). More so for men's wages which
     rise faster.
   - empirically: (among married couples) 
     - high-skill occupation women have children later. Higher-earning men have 
     them earlier.

- Caucutt et al. 2001
  - high waged and high educated women (of recent cohorts) more likely to
    delay childbirth

- Cohen et al. 2013
  - analyse effect of Israeli subsidy
  - along the way, find that income affects fertility negatively at low income,
    positively at high income, supporting Becker and Tomes 76

- Becker and Tomes 1976
  - investment in children quality and quantity when quality must be equal
    across children; and some quality comes from an innate endowment.
  - "observed" quantity-of-children income elasticity rises as income rises
  - while "observed" quality-of-children elasticity falls as income rises, 
    b/c more of total quality must be paid for by parents themselves. 
  - the need to pay more for quality results in lower quantity, since total cost is 
    quantity X quality
  - something fishy about argument somehow...

- Lindo 2008
  - men who lose a job have children sooner, but fewer children in total
  - controls for all skills effects
  - so overall effect of income is positive, but timing changes

- Monstad et al. 2008
  - using an education reform to vary education
  - no evidence of a causal effect on fertility, but births start later
  * maybe the usual correlation reflects underlying skills. Your data could
    test this.

- Autor et al. 2017
  - the China shock (exogenous shock to income) 
  - they use Chinese import penetration by industry (instrumented using
    other countries) times industry employment shares; and separate it for
    males and females
  - shocks to male employment decreased fertility and marriage; 
    increased unmarried mothers
  - shocks to female employment *increase* marriage and motherhood
  * this roughly seems to be the consensus
  * Whereas, you find that male and female *skills* (as 'instrumented' by PGS)
    have the same effect on fertility - i.e. less skills = higher fertility

# WHAT MODEL FITS YOUR DATA
## ASSUMPTIONS
* Genetics produce human capital h. 
* h and education are compliments in producing earnings
* There are income and substitution effects on fertility:
  - income effects are stronger at high income levels (Becker and Tomes 1976)
    and among couples
* People seek partners when in the labour market. Women match a person, observe
  their quality, and decide whether to have a first child.
* Then they (may) have more children, either with the partner or not 
  (high quality partners stay around more)
* High quality partners increase household income

## RESULTS
* People with low h are more likely to leave education early and enter the labour market
  - because e and h are complements
  - this is part of the substitution effect, as they expect to spend some time
    raising children rather than being in work
* At low values of h, increases in h reduce fertility (substitution effect dominates)
  at higher values, smaller reductions or increases (income effect stronger)
  - because h predicts income
* In 1-parent households, increases in h reduce fertility (sub. effect), in 
  2-parent households, smaller reductions or increases (income effect stronger)
  - because h predicts income and because stronger income effect among couples
* People with low h start having children earlier (b/c leave educ earlier)
  - so part of the substitution effect comes via this
  - controlling for this the income effect dominates
  - hence increases in h reduce fertility less once we control for AFLB

## THOUGHTS
* Should we distinguish between men and women in couples?
  - within couples, male h should increase fertility (via earnings)
  - female h should decrease it (ditto?)
  - see explore-sex-vs-pay.R
"
```

@@

```{r plot-means-over-time, fig.height = 5.5, fig.cap = "Mean polygenic scores by birth year in UK Biobank. Points are means for 5-year intervals. Lines are 95% confidence intervals. Green lines show a significant linear increase over time (p < 0.05/33). Red lines show a significant decrease."}

@@

pgs_over_time %>% 
      mutate(score_name = pretty_names(score_name)) %>% 
      ggplot(aes(YOB, group = score_name, colour = Change)) + 
      geom_pointrange(aes(ymin = score_lo, y = score, ymax = score_hi), 
        fatten = 1.25) +
      facet_wrap(~score_name, ncol = 5) +
      my_hline +
      scale_color_manual(values = c("-" = "red", "+" = "darkgreen", "o" = "black")) +
      theme(
        strip.text  = element_text(size = 6), 
        axis.text.x = element_text(size = 7),
        legend.position = "none"
      ) +
      labs(y = "PGS", x = "Birth year (5 year intervals)")

@@

Why do selection effects have *opposite* signs across older and younger mothers
at first live birth, and opposite signs when AFLB is controlled for? A possible
answer is given by the economic theory of fertility [@becker1960economic; @willis1973new].
According to this, increases in potential earned income affect fertility via two
opposing channels. There is an "income effect" by which children become more
affordable, like any other good. There is also a "substitution effect": since
childrearing has a cost in time, the opportunity cost of childrearing increases
if one's market wage is higher. The income effect would lead higher earners to
have more children. The substitution effect would lead higher earners to have
fewer children. It is often assumed that the substitution effect will dominate
for lone parents, or those in unstable relationships, since they have less
opportunity to share childcare responsibilities, while the income effect will
dominate for couples who are able to reap the gains from specialization.

@@

# Discussion

@@

An alternative theory is that traits selected for are linked to
externalizing behaviour and low time discount rates, via the channel of
early sexual behaviour [@Mills2020.05.06.081273]. The data here provide
some support for this: scores which might plausibly be linked to
externalizing behaviour, like ADHD and age at smoking initiation, are
selected for. However, this theory is less good at explaining variation
in selection across the full range of scores, including physical measures, e.g.
waist-hip ratio and BMI. Externalizing behaviour also does not explain
why selection effects change sign among older parents. Overall, we believe
that the economic theory is the leading explanation.

PGS which correlate with high (low) earnings and more (less) education are being
selected against (selected for). In addition, many of the phenotypes under
positive selection are linked to poor health, or would be seen as undesirable to
have by many people. For example, most people would probably prefer to have high
educational attainment, a low risk of major depressive disorder, and a low risk
of coronary artery disease, but natural selection is pushing against these
traits. Potentially, this could increase the health burden on modern societies,
but that depends on effect sizes. Our results suggest that naïve estimates can
be affected by sample ascertainment bias. This problem may be less serious in
surveys which aim to be representative (as the UK Biobank does not).
Nevertheless, not all respondents consent to the collection of genetic data: for
instance, completion rates for genotype data in the US Health and Retirement
Study were around 80-85% [@hrs2020]. Researchers should be aware of the risks
of ascertainment when studying modern natural selection.

@@

Research on natural selection is controversial. It brings to mind eugenics,
which has been used to justify mass murder and human rights violations, in the
past and possibly the present [@mao1998chinese; @sleeboom2011genetic;
@zenz2020]. Whilst we claim scientific legitimacy for our empirical results, we
have no special authority to speak on how society should react to
them. This can only be resolved by open debate, in which scientists take part as
equals among many. We therefore choose not to discuss policy implications here.

@@

Tables \@ref(tab:tbl-siblings-by-period) and
\@ref(tab:tbl-children-by-period) summarize the results. There is no
change in natural selection within the parents' generation. In the
respondents' generation, effect sizes were significantly larger in absolute size among the late-born for five PGS: autism, cognitive ability, EA2, EA3 and
extraversion. These changes are inconsistent with the intergenerational change, 
where effect sizes were if anything larger among the earlier, parents' generation.
Three PGS showed significant changes in sign: alcohol use
(positive to negative); conscientiousness (negative to positive); and
type II diabetes (negative to positive). Overall, there is weak evidence for
change over time, and the clearest result is that the direction of selection is
broadly consistent.

@@

@@

Another way to examine the "earnings" theory of natural selection is to
compare selection effects of PGS with their genetic correlation with
educational attainment (EA3). Since EA3 strongly predicts earnings, if
earnings drives differences in fertility, we'd expect a correlation
between the two sets of results. Figure
\@ref(fig:plot-rgs-by-effect-size) shows this is so.
